---
title: "RnaSeq pipeline"
output:
  github_document:
    html_preview: true
    toc: true
    toc_depth: 4
---

## Getting Started

1. Create a new folder in fs3 or any other location with the project name - called the "project folder" in this README.   
2. Create a bin/ folder in the project folder.   
3. Download and copy the scripts to the bin/ folder.   

## Running the pipeline

### 1. Analysis info file
A central part of the pipeline is the **analysis info** file. It has information about the project location, the original run folder, the reference fasta, gtf and bed files, and the parameters used throughout the analysis.   

#### Format of the analysis info file
The analysis info file is a simple .txt file, with each line providing information. Parameters are separated by the semicolon (i.e ";") character.   

The following is an example of the analysis info file:   

--------------------------------------------------------------------------------------------------   
**working_directory =** /mnt/cgs-fs3/Sequencing/Pipelines/RNASeq_pipeline/example_small_files/    
**run_folder =** /mnt/cgs-fs3/Sequencing/Pipelines/RNASeq_pipeline/example_small_files/xxx   
**run_samplesheet =** /mnt/cgs-fs3/Sequencing/Pipelines/RNASeq_pipeline/example_small_files/xxx    
**bcl2fastq_output =** /mnt/cgs-fs3/Sequencing/Pipelines/RNASeq_pipeline/example_small_files/fastq/   
**readType =** pairedEnd    
**reference_genome =**    
**trimgalore_params =** --gzip; --paired; --fastqc; --fastqc_args '--nogroup --extract'   
**mapping_params =** --runThreadN 4; --outSAMtype BAM SortedByCoordinate; readFilesCommand zcat   
**ncores =** 8   
**htseq_params =** -a 10; -m union; -s reverse; -t exon   
**strand = ** reverse   
--------------------------------------------------------------------------------------------------   

The following is the explanation of the analysis info file:   

--------------------------------------------------------------------------------------------------   
**working_directory =** *\<path to directory of the analysis\>*   
**run_folder =** *\<path to the run folder\>*    
**run_samplesheet =** *\<sample sheet to be used to generate fastq files. This is created using the Illumina Expert Manager\>*    
**bcl2fastq_output =** *\<path to output of bcl2fastq. The defaults is fastq/ and the folder will be created automatically\>*    
**readType =** *\<either pairedEnd or singleEnd\>*     
**reference_genome =** *\<path to the STAR reference genome folder that will be used at the mapping step\>*   
**trimgalore_params =** *\<parameters to be passed to trim galore\>* 
**mapping_params =** *\<parameters to be passed to star\>*
**ncores =** *\<Number of cores to use to pararellize analysis\>*    
**htseq_params =** *\<parameters to be passed to htseq-count\>*
**strand =** *\< expected mapping strand \>*   
--------------------------------------------------------------------------------------------------    
   
<br>   

#### Reference genome   
The reference genome should be the output of STAR's *genomeGenerate* command. We currently download genomes from [ensembl](ftp://ftp.ensembl.org/pub/).      

#### How to create the analysis info file

Use the bin/analysis_info.txt.   

```{r, engine = 'bash', eval = FALSE}
$ python bin/analysis_info.py
```

This will create a file analysis_info.txt, which you can open in a text editor and fill.

<br>   

### 2. Obtain FASTQ files

#### Using bcl2fastq
**Script:** bin/run_bcl2fastq.py   

**Input:** the analysis_info.txt file, which has information about the **run folder**, the **run samplesheet**, and the **bcl2fastq_output** folder. If you have more than one run, you can run this command for each run by changing the run\_folder and run\_samplesheet fields in the analysis\_info.txt file).   

**Command:**   
```{r, engine = 'bash', eval = FALSE}
$ python bin/run_bcl2fastq.py --analysis_info_file analysis_info.py   
```   

**Output:** fastq/ folder, which has a project subfolder with the fastq.gz files, among other files and subfolders.  


#### Downloading data from basespace (Currently used for Lexogen projects)

For lexogen projects, we currently download the data from Basespace. Normally, the downloaded data will have the following structure:

-   A main folder with the project name
-   Subfolders, which correspond to each of the sequenced samples
-   Within each subfolder, four fastq.gz files from the corresponding sample

The tasks done are:

-   Unzip the fastq files:
    -   Move to the downloaded folder.
    -   Use the following commands:

```{r, engine = 'bash', eval = FALSE}
$ ls -d * | parallel -j 4 --no-notice "cd {} ; gzip -d *"      
```

-   Change the folder name of each sample to a string consisting of **the sample name, an underscore, a capital S, and one or more numbers (eg. sample1\_S1)**. You can use the following commands, however **make sure that the sample names consists of the original sample name, an underscore, a capital S, and one or more numbers (eg. sample1\_S1)**:
    -   Check what the outcome would be. This command will print the original folder name, followed by a space and the new folder name:
```{r, engine = 'bash', eval = FALSE}
$ for i in $(ls -d *); do sample_name=$(ls ${i} | sed 's/.*\///g' | sed 's/_L00[[:digit:]]\+.*//g'  | sort | uniq); echo ${i} ${sample_name}; done         
```           
    -   If it looks correct, use mv to change the folder names:    
```{r, engine = 'bash', eval = FALSE}     
$ for i in $(ls -d *); do sample_name=$(ls ${i} | sed 's/.*\///g' | sed 's/_L00[[:digit:]]\+.*//g'  | sort | uniq); mv ${i} ${sample_name}; done     
```       

-   For each sample, concatenate the four files into one:    
    -   The following command tests whether the outcome of the command is right. It prints the sample name, it's four fasta files, and the name of the output concatenated file.    
```{r, engine = 'bash', eval = FALSE}
$ ls -d * | parallel -j 4 --no-notice "echo {}; ls {}/{}*L001_R1*  {}/{}*L002_R1* {}/{}*L003_R1*  {}/{}*L004_R1*; echo {}/{}_R1_001.fastq"     
```    
    -   If that looks correct, concatenate the files for R1 reads:   
```{r, engine = 'bash', eval = FALSE}
$ ls -d * | parallel -j 4 --no-notice "cat {}/{}*L001_R1*  {}/{}*L002_R1* {}/{}*L003_R1*  {}/{}*L004_R1* > {}/{}_R1_001.fastq"     
```    
    -   And for R2 reads for paired end reads:    
```{r, engine = 'bash', eval = FALSE}
$ ls -d * | parallel -j 4 --no-notice "cat {}/{}*L001_R2*  {}/{}*L002_R2* {}/{}*L003_R2*  {}/{}*L004_R2* > {}/{}_R2_001.fastq"    
```    

<br>
    
### 3. Move the reads to a new folder named rawReads    
Use mv command in bash.   

<br>   

### 4. Create sample names file   
**Bash command:**      
```{r, engine = 'bash', eval = FALSE}
$ ls rawReads | sed 's/_R1_001.fastq//g' | sed 's/_R2_001.fastq//g' | sort | uniq > sample_names.txt   
```

<br>   

### 5. Quality control of Fastq files
To run fastqc in all the samples, use the script bin/qcReads.py.

**Script:** bin/qcReads.py

**Arguments:**   

--------------------------------------------------------------------------------------------------
-h, --help	show this help message and exit
-v, --version	show program's version number and exit
--analysis_info_file	ANALYSIS_INFO_FILE. Text file with details of the analysis. Default=analysis_info.txt
--in_dir	IN_DIR. Path to folder containing fastq files. Default=rawReads/
--out_dir	OUT_DIR. Path to out put folder. Default=rawReads/
--sample_names_file	SAMPLE_NAMES_FILE. Text file with sample names to run. Default=sample_names.txt
--------------------------------------------------------------------------------------------------

**Output:**
Fastqc files and folders will be created for each sample in the rawReads/ folder

**Command example:**

```{r, engine = 'bash', eval = FALSE}
$ python bin/qcReads.py --analysis_info_file analysis_info.txt --in_dir rawReads/ --out_dir rawReads/ --sample_names_file sample_names.txt
```

<br>   

### 6. Table and plot of number of reads per sample

**Script:** bin/indexQC.R

**Requires:** ggplot2 and reshape R libraries

**Input:** The input directory should have the *fastqc\_data.txt* files of all samples of the project. The 7th line in the fastqc\_data.txt files have the total number of sequences. This is the information that the script collects.

**Command:**   

```{r, engine = 'bash', eval = FALSE}
$ /usr/bin/Rscript bin/indexQX.R <input dir> <output dir>
```

**Command example:**   

```{r, engine = 'bash', eval = FALSE}
$ /usr/bin/Rscript bin/indexQX.R rawReads/ Report/figure/data/
```

**Output:** The output directory will have a csv file with the Total number of Sequences and a bar plot with the proportion of reads represented by each sample from the total number of reads from the run.

<br>   

### 7. FastQC plots 

**main script:** bin/fastqc_tables_and_plots.py    

**sub scripts:** bin/create_fastqcPlots_allSamples.R; bin/create_fastqcPlots_perSample.R; bin/create_fastqcTables.py   

**Arguments main script:**     

--------------------------------------------------------------------------------------------------
-h, --help   show this help message and exit
-v, --version        show program's version number and exit
--in_dir IN_DIR       Path to folder containing fastq files. Default=rawReads/
--out_dir_report OUT_DIR_REPORT Path to out put folder. Default=Report/figure
--suffix_name SUFFIX_NAME Suffix to optionally put to the output name. Default=\'\'
--sample_names_file SAMPLE_NAMES_FILE Text file with sample names. Default=sample_names.txt
--plot_device PLOT_DEVICE Specify the format of the plot output. Default=png
--------------------------------------------------------------------------------------------------


**Command example:**      
```{r, engine = 'bash', eval = FALSE}
$ python bin/fastqc_tables_and_plots.py --in_dir rawReads/ --out_dir_report Report/figure/rawQC/ --suffix_name _raw --sample_names_file sample_names.txt 
```

<br>   

### 8. Trim low quality bases and adapters   

**main script:** bin/trimmingReads.py   

**sub script:** bin/trimming_summary.R     

**analysis info file:** Define the trimgalore parameters that you want to pass to trim galore and the number of cores.      

**Lexogen projects analysis info: ** For lexogen projects, the argument *--clip_R1 12* is added in the analysis_info.txt file (trimgalore_params line) to remove the first 12 bases as recommended. 

**Arguments:**   

--------------------------------------------------------------------------------------------------
-h, --help   show this help message and exit
-v, --version        show program's version number and exit
--analysis_info_file ANALYSIS_INFO_FILE. Text file with details of the analysis. Default=analysis_info.txt
--in_dir     IN_DIR. Path to folder containing fastq files. Default=rawReads/
--out_dir    OUT_DIR. Path to out put folder. Default=trimmedReads/
--sample_names_file  SAMPLE_NAMES_FILE. Text file with sample names to run. Default=sample_names.txt
--------------------------------------------------------------------------------------------------

**Command example:**      
```{r, engine = 'bash', eval = FALSE}
$ python bin/trimmingReads.py
```

<br>   

### 9. Trimming summary   

The script bin/trimming_summary creates a table with the number of raw sequences, the number of sequences after trimming, and the percentage of sequences removed after trimming.

It expects an input directory of raw data with a fastqc output file for each sample (fastqc_data.txt).   

It also expects an input directory or trimmed data with a fastqc output file for each sample (fastqc_data.txt).   
   
**Arguments:**   
```{r, engine = 'bash', eval = FALSE}
/usr/bin/Rscript bin/trimming_summary.R <raw_data_indir> <trimmed_data_indir> <outdir>   
```   

**Command example:**   
```{r, engine = 'bash', eval = FALSE}
/usr/bin/Rscript bin/trimming_summary.R rawReads/ trimmedReads/ Report/figure/data/
```

**Output:** A csv table in the output directory.   
  

<br>   

### 10. Trimming QC plots   

**Make sure the arguments, specially *in_dir* and *out_dir* arguments, correspond to the trimmed reads corresponding folders**

**main script:** bin/fastqc_tables_and_plots.py   

**sub scripts:** bin/create_fastqcPlots_allSamples.R; bin/create_fastqcPlots_perSample.R; bin/create_fastqcTables.py

**Arguments main script:**     

--------------------------------------------------------------------------------------------------
-h, --help   show this help message and exit
-v, --version        show program's version number and exit
--in_dir IN_DIR       Path to folder containing fastq files. Default=rawReads/
--out_dir_report OUT_DIR_REPORT Path to out put folder. Default=Report/figure
--suffix_name SUFFIX_NAME Suffix to optionally put to the output name. Default=\'\'
--sample_names_file SAMPLE_NAMES_FILE Text file with sample names. Default=sample_names.txt
--plot_device PLOT_DEVICE Specify the format of the plot output. Default=png
--------------------------------------------------------------------------------------------------

**Command example:**  
```{r, engine = 'bash', eval = FALSE}
$ python bin/fastqc_tables_and_plots.py --in_dir trimmedReads --out_dir_report Report/figure/trimmedQC/ --suffix_name _trimmed --sample_names_file sample_names.txt
```

<br>   

### 10. (Optional for Lexogen) Trim polyA sequences   

PolyA sequences may also affect the mapping of reads if the alignment tool used involves alignment of reads from end-to-end. The tool used for mapping in this analysis is STAR, as explained in the following section.

STAR does local alignment and clipping of 3’-end of reads in case that such an alignment gives a better score. This means that the majority of polyA sequences are clipped from reads during the alignment step. For this reason it is not strictly necessary to trim polyA sequences from the reads. 

To remove polyA sequences I have used prinseq (argument -trim_right 8, but may change). 

<br>


### 11. Mapping   

**main script:** mappingReads.py   

**software:** samtools, STAR

**Define the reference genome in the analysis_info.txt**   

** Define the number of cores to be used in analysis_info.txt, normally 2 works fine**   

**Comments:** a temp dir can be a local directory. Transfer of data between servers (i.e fs3 and cluster) is often very slow.


**Arguments main script:**   

--------------------------------------------------------------------------------------------------
-h, --help   show this help message and exit
-v, --version        show program's version number and exit
--in_dir IN_DIR       Path to folder containing fastq files. Default=trimmedReads/
--out_dir OUT_DIR     Path to out put folder. Default=alignedReads/
--temp_dir TEMP_DIR   Path to a temp dir folder. Default=none
--sample_names_file SAMPLE_NAMES_FILE. Text file with sample names to map. Default=sample_names_info.txt
--------------------------------------------------------------------------------------------------

**Command example:**   
```bash
$ python bin/mappingReads.py --temp_dir ~/temp_mapping
```

<br>


### 12. Mapping QC     

#### 12a. Mapping summary    

**main script:** bin/mappingQC.py    

**other scripts:** bin/mapping_summary.R   

**Input:** folder of mapped reads with STAR output files *Log.final.out   

**Command example:**         
```{r, engine = 'bash', eval = FALSE}
python bin/mappingQC.py --run mapping_summary --in_dir alignedReads --out_dir_report Report/figure/mappingQC/
```

**Ouput:** csv file in the out_dir_report with percentage of uniquely mapped reads, reads mapped to multiple loci, unmapped reads, starting number of reads, and number of mapped and usable reads (uniquely mapped reads)

<br>


#### 12b. Gene body coverage   

**main script:** bin/mappingQC.py

**main tools:** /usr/local/minicondaexport/bin/geneBody_coverage.py

**Input:**  folder of mapped reads with bam files.

**Bed file:** bedFile_10k specified in the analyis info file.    

**Command example:**
```{r, engine = 'bash', eval = FALSE}
python bin/mappingQC.py --run gene_body_coverage --in_dir alignedReads --out_dir alignedReads/QC --out_dir_report Report/figure/mappingQC/
```   

**Output:** pdf file with gene body coverage plot in the out_dir_report. *I re-do this plot using the 10KGenes.geneBodyCoverage.r, which is also an output from the script, and is stored in the out_dir folder*

<br>


#### 12c. Genomic context of mapped reads.   

**main script:** bin/mappingQC.py

**main tools:** /usr/local/picard/picard.jar CollectRnaSeqMetrics       

**other scripts:** bin/read_distribution_genomic_context.R, bin/mapping_distribution.R   

**Input:** folder of mapped reads with bam files.   

**Other files:** ref flat file and ribosomal rna interval list, both specified in the analysis info file.    

**Other options:** *strand*, specified in the analysis info file.    

**Command example:**
```{r, engine = 'bash', eval = FALSE}
python bin/mappingQC.py --run picard_tools --in_dir alignedReads --out_dir alignedReads/QC --out_dir_report Report/figure/mappingQC/
```

**Output:** csv and txt files in out_dir. read_distribution_genomic_context png file and strand_mappingQC csv file in out_dir_report.   
   
<br>   
   
### 13. Counting of reads

**main script:** bin/countingQC.py   

**software:** samtools and htseq-count

**gtf_file:** gtf file specified in the analysis info file  

**Other options:** *htseq_params*, specified in the analysis info file

**Command example:**
```{r, engine = 'bash', eval = FALSE}
python bin/countingReads.py --in_dir alignedReads --out_dir countedReads 
```   

**Output:** *_count.txt* files for each samples with counts for each gene.   

### 14. Counting QC

#### 14a. Counting QC part1

**Script:** bin/countingQC_part1.R   

**R packages required:** RColorBrewer, matrixStats, gplots, ggplot2, reshape, GenomicRanges, rtracklayer, Rsamtools, grid.   

**Input:** folder with _count.txt files

**Other files (optional):** mapping_summary.csv, output of mappinQC  

**Command: **   
```{r, engine = 'bash', eval = FALSE}
/usr/bin/Rscript bin/countingQC_part1.R <indir> <outdir> <mapping summary csv>  
```

**Command example: **      
```{r, engine = 'bash', eval = FALSE}
/usr/bin/Rscript bin/countingQC_part1.R countedReads/ Report/figure/countingQC Report/figure/mappingQC/mapping_summary.csv
```

**Output:** files and plots in outdir

#### 14b. Counting QC part2

**Script: ** bin/countingQC_part2.R   

**R packages required: ** gplots, rtracklayer, reshape, ggplot2.   

**Input: ** folder with _count.txt files

**Other files: ** GTF file

**Command: **
```{r, engine = 'bash', eval = FALSE}
/usr/bin/Rscript bin/countingQC_part2.R <indir> <outdir> <gtf file>
```

**Command example: **
```{r, engine = 'bash', eval = FALSE}
/usr/bin/Rscript bin/countingQC_part2.R countedReads/ Report/figure/countingQC/ /mnt/cgs-fs3/Sequencing/Genome/Mouse/gtf/ensembl/grcm38/release-84/Mus_musculus.GRCm38.84.gtf
```

**Output:** files and plots in outdir   
 
 
### 15. EdgeR and DESeq2   
At the moment, edgeR is used to do the differential expression analysis. DESeq2 is only used to do PCA plots with **rlog** transformed data.      

#### 15a. Create requiered files:   
**Scripts: ** bin/deseq2_arguments.py, bin/edger_arguments.py      

**Commands: **   
```{r, engine = 'bash', eval = FALSE}
python bin/edger_arguments.py   
python bin/deseq2_arguments.py
```

**Output: **   
- **sample_info.csv.** This file will have columns **SampleID** and **Group**. You can add a **Sibship** column for paired analysis. Fill this file manually.    
- **comparisons.csv.** This file will have columns **baselineGroup** and **comparisonGroup**. The baselineGroup is normally the WT or control group. List the pairwise comparisons you want to make, using the corresponding group names.   
- **deseq2_arguments.txt.** Fill this file manually. Explanation and example below.    
- **edger_arguments.** Fill this file manually. Explanation and example below.   


**deseq2 arguments txt file explanation:**   

---------------------------------------------------------------------------------------------------
indir = *\<folder with _count.txt files\>*
outdir = *\<output dir for results\>*
sample_info = *\<csv file with columns SampleID and Group (and Sibship for paired analysis)\>*
comparisons = *\<csv file with columns baselineGroup and comparisonGroup with a list of comparisons\>*
design = *\<either pairedSamples OR non-pairedSamples\>*
gtfFile = *\<gtf file with gene info. Can be copied from the analysis_info file\>*
---------------------------------------------------------------------------------------------------


**deseq2 arguments txt file example:**   

---------------------------------------------------------------------------------------------------
indir = countedReads/   
outdir = deseq2/   
sample_info = sample_info.csv   
comparisons = comparisons.csv   
design = non-pairedSamples   
gtfFile = /mnt/cgs-fs3/Sequencing/Genome/Mouse/gtf/ensembl/grcm38/release-84/Mus_musculus.GRCm38.84.gtf   
---------------------------------------------------------------------------------------------------


**edger arguments txt file explanation:**   

---------------------------------------------------------------------------------------------------
indir = *\<folder with _count.txt files\>*
outdir = *\<output dir for results\>*
sample_info = *\<csv file with columns SampleID and Group (and Sibship for paired analysis)\>*
comparisons = *\<csv file with columns baselineGroup and comparisonGroup with a list of comparisons\>*
min.cpm = *\<Filtering threshold: miminum count per million value a gene should have to be kept\>*
min.nsamples = *\<Filtering threshold: miminum number of samples with cpm specified above a gene should have to be kept\>*
design = *\<either pairedSamples OR non-pairedSamples\>*
gtfFile = *\<gtf file with gene info. Can be copied from the analysis_info file\>*
---------------------------------------------------------------------------------------------------


**edger arguments txt file example:**   

---------------------------------------------------------------------------------------------------
indir = countedReads/    
outdir = <edger/    
sample_info = sample_info.csv   
comparisons = comparisons.csv   
min.cpm = 1   
min.nsamples = 1   
design = non-pairedSamples   
gtfFile = /mnt/cgs-fs3/Sequencing/Genome/Mouse/gtf/ensembl/grcm38/release-84/Mus_musculus.GRCm38.84.gtf   
---------------------------------------------------------------------------------------------------

   
#### 15b. DESeq2 PCA plot   

**main script: ** bin/deseq2.R   

**other scripts: ** bin/deseq2_functions.R   

**R packages: ** DESeq2, gplots, rtracklayer, ggplot2, gridExtra, matrixStats, dplyr.   

The following script uses DESeq2 package to transform raw counts to rlog values.   

The input directory, output directory, and other arguments need to be specified in the deseq2_arguments.txt.       
**Command: **   
```{r, engine = 'bash', eval = FALSE}   
/usr/bin/Rscript bin/deseq2.R deseq2_arguments.txt   
```   

Output: PCA and heatmap plots in the outdir specified.   


#### 15c. Differential gene expression.   

**Command: **   
```{r, engine = 'bash', eval = FALSE}
/usr/bin/Rscript bin/edger.R edger_arguments.txt
```



 







